Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-22 paper.qmd
spell check

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---
# Current Todo List
This is a list of tasks that need to be completed before I consider the paper complete. Please keep this in mind when reading.

@@

Uncertainty visualisation is a somewhat new subfield of visualisation, with early papers that specifically reference an "uncertainty visualisation" popping up in the late 90s and early 2000s (*Cite Gap 1*). Despite the youth of the field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves (*Cite Gap 2*). Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satelite images (*Cite Gap 1*). In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts and is no longer bound by geospatial information (*Cite Gap 1*).

@@

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field. @Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of each uncertainty visualisation paper.

There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature. Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016]. Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017]. Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These difficulties in noisy uncertainty visualisation findings arises reguardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].

This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The absence of an econompasing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

@@

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.

@@

## Taxonomy definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

When considering the use of taxonomies to define something inherrently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:

@@

@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field and in doing so, highlights an interesting connection between inference and uncertainty. Descriptive statistics simply describe our sample as it is and summarises large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case. Around the 19th century in England, *posivitism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023]. By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect useage. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e. that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023]. This subtle shift, from descriptive statistics to inferential statistics was historically shunned *due to the fact it introduced the unknowable*, or in other words, uncertainty.

This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagreee with that statement. Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it, and since probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty. This is not necessarily true, and related distinction was made by @Spiegelhalter2017  commented on a differentiation between precise random events (such as the probability of a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). A sample variance is not unknown, and therefore it is not uncertain, rather it is a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty, but in isolation it is not uncertain.

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

@@

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty. This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of understanding and quantifying them are ad-hoc at best. @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy. None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication. The need for a mathematical definition of this field is put simply by Freeman Dyson in his fameous Birds and Frogs speech:

@@

In order to visualise something, it needs to be quantifiable. In order to quanitfy something, it needs to be mathematically defined. The broad classes of what is consdiered "uncertainty" is not currently quantifiable in any way that is not ad-hoc. The only apsects of uncertaitny that are currently quantifiable are confidence intervals, prediction intervals, and related terms. Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis. Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.

## Mistakes made when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence). When uncertainty visualisaiton authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself. In these cases, the authors are more confused about what they are doing than the participants.

@@

@Hofman2020 performed an experiment where some participants were shown a sampling distribution of the mean, and others were shown a prediction interval around the mean, for the effectiveness of a particular treatment. The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval. A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectivness of the treatment*. It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017]. The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* statistics. This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.

@@

Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem. @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant. @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results. This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values. The surprise that both bias and variance contribute to final uncertainty shows that authors do not understand both will contribute to the *distance* of your estimate to the value you are drawing inference on (AKA error).

The misunderstanding that descriptive statistics also extends to visualisation. The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature. @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Exploratory visualisations often differ from descriptive statistics because the explicit statistic we are drawing inference on is less explicit. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which may have its own associated uncertainty. @Hullman2021 argue that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference. While we agree with this sentiment, it is clear that uncertainty *cannot* be defined without a specific motivating question, and therefore trying to include uncertainty in an exploratory visualisation stage (which by definition does not have a hypothesis) is not possible.

This confusion (or indifference) also creates many imprecisely named papers. Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*). While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.

@@

There are a small handful of taxonomies that highlight some interesting features in the way uncertainty is visualised. @Kinkeldey2014 categorised uncertainty according to whether or not it was:
1. Implicit (a sample) or explicit (a depiction of mass)
2. Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty)
3. visually integral or separable (the uncertainty can be separated from the data and read independently) 
4. coincidence (uncertainty and data are represented in the same plot) or adjacent
5. static or dynamic (animated, interactive ect)
Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014]. This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution). A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally aline with the intrinsic/extrinsic distinction. The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two". This consideration will become more relevant later in this review.

@@

Most of these taxonomies are created by observing existing uncertainty visualisations, which means they suffer from many of the same conceptual issues that the field has with uncertainty in general. Additionally, there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation. The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical. @Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general. The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model. Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, this is something we will show in the following sections of this paper.

### Contextual Information
One of the key components of uncertainty and inference is the noise and signal dichotomy. When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lense of a Bernoulli distribution that we can view these two things as the same. The fact is that this boiling down, of a real world event into data, is not something that happens independent of the statisticians, but is just as much as product of our interests (the particular relationship we hope to draw inference on) as it is a product of the event's natural properties. This is expressed simply in the text with an extension to the coin flipping example:

@@

With this in mind, let us turn to the concept of "contextual" information. For the purposes of this review, we will consider contexual information to be anything that makes it easy for the audience to connect the data from a probablistic kind to its real world parallel, but is largely irrelevant in the abstract probability space. A key example of this is spatial information, such as borders, road locations, terrains, etc. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model. @fig-spatial shows the proces of boilling information down into data, visualising it, and including contextual information to improve our scientific reasoning. The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:

@@

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymetry are regularly compared because they are both "uncertainty visualisations". This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results. This issue likely results in the lacking mathematical definition preventing us from defining and quantifying "uncertainty" mathematically and then including this quantification in a well defined framework such as the grammar of graphics.

>>>>

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---

@@

Uncertainty visualisation is a somewhat new subfield of visualisation, with early papers that specifically reference an "uncertainty visualisation" popping up in the late 90s and early 2000s (*Cite Gap 1*). Despite the youth of the field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves (*Cite Gap 2*). Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satellite images (*Cite Gap 1*). In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts and is no longer bound by geospatial information (*Cite Gap 1*).

@@

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field. @Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of each uncertainty visualisation paper.

There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature. Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016]. Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017]. Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].

This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

@@

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.

@@

## Taxonomy definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

When considering the use of taxonomies to define something inherently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:

@@

@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field and in doing so, highlights an interesting connection between inference and uncertainty. Descriptive statistics simply describe our sample as it is and summarises large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case. Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023]. By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect usage. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e. that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023]. This subtle shift, from descriptive statistics to inferential statistics was historically shunned *due to the fact it introduced the unknowable*, or in other words, uncertainty.

This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagree with that statement. Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it, and since probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty. This is not necessarily true, and related distinction was made by @Spiegelhalter2017  commented on a differentiation between precise random events (such as the probability of a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). A sample variance is not unknown, and therefore it is not uncertain, rather it is a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty, but in isolation it is not uncertain.

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failing to do it well can cause downstream effects and damage the effectiveness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

@@

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty. This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of understanding and quantifying them are ad-hoc at best. @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy. None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication. The need for a mathematical definition of this field is put simply by Freeman Dyson in his famous Birds and Frogs speech:

@@

In order to visualise something, it needs to be quantifiable. In order to quantify something, it needs to be mathematically defined. The broad classes of what is considered "uncertainty" is not currently quantifiable in any way that is not ad-hoc. The only aspects of uncertainty that are currently quantifiable are confidence intervals, prediction intervals, and related terms. Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis. Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.

## Mistakes made when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence). When uncertainty visualisation authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself. In these cases, the authors are more confused about what they are doing than the participants.

@@

@Hofman2020 performed an experiment where some participants were shown a sampling distribution of the mean, and others were shown a prediction interval around the mean, for the effectiveness of a particular treatment. The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval. A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*. It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017]. The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* statistics. This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.

@@

Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem. @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modellers express both if they are relevant. @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results. This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values. The surprise that both bias and variance contribute to final uncertainty shows that authors do not understand both will contribute to the *distance* of your estimate to the value you are drawing inference on (AKA error).

The misunderstanding that descriptive statistics also extends to visualisation. The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisation, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature. @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Exploratory visualisations often differ from descriptive statistics because the explicit statistic we are drawing inference on is less explicit. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which may have its own associated uncertainty. @Hullman2021 argue that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference. While we agree with this sentiment, it is clear that uncertainty *cannot* be defined without a specific motivating question, and therefore trying to include uncertainty in an exploratory visualisation stage (which by definition does not have a hypothesis) is not possible.

This confusion (or indifference) also creates many imprecisely named papers. Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*). While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.

@@

There are a small handful of taxonomies that highlight some interesting features in the way uncertainty is visualised. @Kinkeldey2014 categorised uncertainty according to whether or not it was:
1. Implicit (a sample) or explicit (a depiction of mass)
2. Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty)
3. visually integral or separable (the uncertainty can be separated from the data and read independently) 
4. coincidence (uncertainty and data are represented in the same plot) or adjacent
5. static or dynamic (animated, interactive etc)
Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014]. This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution). A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally align with the intrinsic/extrinsic distinction. The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two". This consideration will become more relevant later in this review.

@@

Most of these taxonomies are created by observing existing uncertainty visualisations, which means they suffer from many of the same conceptual issues that the field has with uncertainty in general. Additionally, there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation. The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical. @Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualisation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general. The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model. Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, this is something we will show in the following sections of this paper.

### Contextual Information
One of the key components of uncertainty and inference is the noise and signal dichotomy. When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lens of a Bernoulli distribution that we can view these two things as the same. The fact is that this boiling down, of a real world event into data, is not something that happens independent of the statisticians, but is just as much as product of our interests (the particular relationship we hope to draw inference on) as it is a product of the event's natural properties. This is expressed simply in the text with an extension to the coin flipping example:

@@

With this in mind, let us turn to the concept of "contextual" information. For the purposes of this review, we will consider contextual information to be anything that makes it easy for the audience to connect the data from a probabilistic kind to its real world parallel, but is largely irrelevant in the abstract probability space. A key example of this is spatial information, such as borders, road locations, terrains, etc. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model. @fig-spatial shows the process of boiling information down into data, visualising it, and including contextual information to improve our scientific reasoning. The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:

@@

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations". This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results. This issue likely results in the lacking mathematical definition preventing us from defining and quantifying "uncertainty" mathematically and then including this quantification in a well defined framework such as the grammar of graphics.
